\graphicspath{{notes/3canonical/asy/}}

\setcounter{section}{2}

\section{Canonical Forms}

\subsection{Jordan Forms \& Generalized Eigenvectors}

Essentially everything in these notes is a variation of a single question: given a linear map $\rT$ on a finite-dimensional vector space $V$, how can we find a basis $\beta$ such that the matrix $[\rT]_\beta$ is as close to diagonal as possible? In this chapter we see what is possible when $\rT$ is non-diagonalizable.


\begin{example}{}{jordan1}
	The matrix $A=
	\begin{smatrix}
		-8&4\\
		-25&12
	\end{smatrix}
	\in M_2(\R)$ has characteristic equation
	\[
		p(t)=(-8-t)(12-t)+4\cdot 25=t^2-4t+4=(t-2)^2
	\]
	and thus a single eigenvalue $\lambda=2$ of algebraic multiplicity two. The matrix is non-diagonalizable since the corresponding eigenspace is one-dimensional (Theorem \ref{thm:diagcrit}):
	\[
		E_2 =\cN
		\begin{pmatrix}
			-10&4\\
			-25&10
		\end{pmatrix}
		=\Span\twovec 25
	\]
	However, with respect to a basis $\beta=\{\vv_1,\vv_2\}$ where $\vv_1=\stwovec 25$ is an eigenvector, $[\rL_A]_\beta$ is \emph{upper-triangular} which is better than nothing! How simple can we make this matrix? If $\vv_2=\stwovec xy$, then
	\begin{align*}
		&A\vv_2=\twovec{-8x+4y}{-25x+12y}=2\twovec xy+\twovec{-10x+4y}{-25x+10y}=2\vv_2+(-5x+2y)\vv_1\\
		&\implies [\rL_A]_\beta=
		\begin{pmatrix}
			2&-5x+2y\\
			0&2
		\end{pmatrix}
	\end{align*}
	Since $\vv_2$ cannot be parallel to $\vv_1$, the only thing we cannot have is a diagonal matrix. The next best thing is to make the upper right corner 1; this can be achieved by, for instance, choosing
	\[
		\beta=\{\vv_1,\vv_2\}=\left\{\twovec 25,\twovec 13\right\}
		\implies [\rL_A]_\beta=
		\begin{pmatrix}
			2&1\\
			0&2
		\end{pmatrix}
	\]
\end{example}


\begin{defn}{}{}
	A \emph{Jordan block} is a square matrix of the form
	\[
		J=
		\begin{pmatrix}
			\lambda&1&&\\[-2pt]
			&\lambda&\ddots&\\[-2pt]
			&&\ddots&1\\
			&&&\lambda
		\end{pmatrix}
	\]
	where all non-indicated entries are zero. Any $1\times 1$ matrix is also a Jordan block.\par
	A \emph{Jordan canonical form} is a block-diagonal matrix $\diag(J_1,\ldots,J_m)$ where each $J_k$ is a Jordan block.\par
	A \emph{Jordan canonical basis} for $\rT\in\cL(V)$ is a basis $\beta$ of $V$ such that $[\rT]_\beta$ is a Jordan canonical form.\footnotemark
\end{defn}

\footnotetext{%
	If $\rT=\rL_A$ is left-multiplication by a matrix $A\in M_n(\F)$, we simply refer to a Jordan canonical basis for $A$.}

In Example \ref{ex:jordan1}, $\beta$ is a Jordan canonical basis and $[\rL_A]_\beta$ a Jordan canonical form.\smallbreak

If a map is diagonalizable, then any eigenbasis is Jordan canonical and the corresponding Jordan canonical form is diagonal. What about more generally? Does every non-diagonalizable map have a Jordan canonical basis? If so, how can we find such?


\pagebreak[3]


\begin{example}{}{jordan2}
	We check that
	$\beta=\{\vv_1,\vv_2,\vv_3\}=\left\{\sthreevec 101,\sthreevec 120,\sthreevec 111\right\}$
	is a Jordan canonical basis for 
	\[
		A=
		\begin{pmatrix}
			-1&2&3\\
			-4&5&4\\
			-2&1&4
		\end{pmatrix}
	\]
	Indeed
	\[
		A\vv_1=2\vv_1,\quad 
		A\vv_2=3\vv_2,\quad 
		A\vv_3=\threevec 453=\threevec{1+3}{2+3}{0+3}=\vv_2+3\vv_3\implies [\rL_A]_\beta=
		\begin{pmatrix}
			2&0&0\\
			0&3&1\\
			0&0&3
		\end{pmatrix}
	\]
	This last is a Jordan canonical form for $A$.
\end{example}


\boldsubsubsection{Generalized Eigenvectors}

Example \ref{ex:jordan2} was easy to verify, but how would we go about finding a suitable basis in general? We brute-forced this in Example \ref{ex:jordan1}, but this is not a reasonable way to proceed for harder problems. Considering the shape of a Jordan form, any basis vector corresponding to a column with at most one non-zero entry must be an eigenvector:
\begin{itemize}
  \item In Example \ref{ex:jordan1}, $\vv_1$ is an eigenvector but $\vv_2$ is not.
  \item In Example \ref{ex:jordan2}, $\vv_1$ and $\vv_2$ are eigenvectors but $\vv_3$ is not.
\end{itemize}
The practical question, therefore, is how to fill out a Jordan canonical basis once we have a maximal independent set of eigenvectors. We now define the necessary objects.

\begin{defn}{}{}
	Suppose $\rT\in\cL(V)$ has an eigenvalue $\lambda$. Its \emph{generalized eigenspace} is
	\[
		K_\lambda:=
		\big\{\vx\in V:(\rT-\lambda \I)^k(\vx)=\V0\text{ for some }k\in\N\big\}
		=\smash[b]{\bigcup_{k\in\N}\cN(\rT-\lambda\I)^k}
	\]
	A \emph{generalized eigenvector} is any non-zero $\vv\in K_\lambda$.
\end{defn}

As with eigenspaces, the generalized eigenspaces of $A\in M_n(\F)$ are those of the map $\rL_A\in\cL(\F^n)$.\smallbreak

\begin{examples*}{cont.}{}
	Our earlier Jordan canonical bases consist of generalized eigenvectors.
	\begin{description}
	  \item[{Example \ref{ex:jordan1}}.] We have one eigenvalue $\lambda=2$. Since $(A-2I)^2=
	  \begin{smatrix}
	  	0&0\\
	  	0&0
	  \end{smatrix}$
	  is the zero matrix, every non-zero vector is a generalized eigenvector: that is, $K_2=\R^2$.
	
	  \item[Example \ref{ex:jordan2}.] We see that
	  \[
	  	(A-2I)\vv_1=\V0,\qquad 
	  	(A-3I)\vv_2=\V0,\qquad 
	  	(A-3I)^2\vv_3=(A-3I)\vv_2=\V0
	  \]
	  whence $\beta$ is a basis of generalized eigenvectors. Indeed
	  \[
	  	K_3=\Span\{\vv_1,\vv_2\},\quad 
	  	K_2=E_2=\Span\{\vv_3\}
	  \]
	  though verifying this with current technology is a little awkward\ldots
	\end{description}
\end{examples*}


\pagebreak[3]


To  help compute generalized eigenspaces, we invoke the main result of this section; the proof is postponed due to its meatiness. Make sure to compare this with the diagonalizability criterion (Theorem \ref{thm:diagcrit2}).

\begin{thm}{}{generalized1}
	Suppose that the characteristic polynomial of $\rT\in\cL(V)$ splits over $\F$:
	\[
		p(t)=(\lambda_1-t)^{m_1}\cdots (\lambda_k-t)^{m_k}
	\]
	where the $m_j$ are the algebraic multiplicities of the \emph{distinct} eigenvalues $\lambda_j$. Then:
	\begin{enumerate}\itemsep0pt
	  \item For each eigenvalue, \ (a) \ $K_{\lambda}=\cN(\rT-\lambda \I)^{m}$, \ and \ (b) \ $\dim K_{\lambda}=m$.
		\item $V=K_{\lambda_1}\oplus\cdots\oplus K_{\lambda_k}$. In particular, there exists a basis of generalized eigenvectors.
	\end{enumerate}
\end{thm}

With regard to part 2, we shall eventually be able to \emph{choose} the basis to be Jordan canonical. In conclusion: a map has a Jordan canonical basis if and only if its characteristic polynomial splits.



\begin{examples}{}{}
	We revisit the previous example in this new language, and consider a second matrix from scratch.
	\begin{enumerate}
  	\item (Example \ref{ex:jordan2}) \ $A=
  	\begin{smatrix}
			-1&2&3\\
			-4&5&4\\
			-2&1&4
		\end{smatrix}$
		has characteristic polynomial $p(t)=(2-t)^{\textcolor{blue}{1}}(3-t)^{\textcolor{red}{2}}$. Therefore,
		\begin{gather*}
			K_2=\cN(A-2I)^{\textcolor{blue}{1}}=\Span\threevec 101\implies\dim K_2={\textcolor{blue}{1}}\\
			K_3=\cN(A-3I)^{\textcolor{red}{2}}=\cN\begin{pmatrix}
			2&-1&-1\\0&0&0\\2&-1&-1
			\end{pmatrix} =\Span\left\{\threevec 120,\threevec 111\right\}\implies\dim K_3={\textcolor{red}{2}}
		\end{gather*}
		In accordance with the theorem, $\R^3=\textcolor{blue}{K_2}\oplus \textcolor{red}{K_3}$.
	
	  \item We find the generalized eigenspaces of the matrix $B=
	  \begin{smatrix}
			5&2&-1\\
			0&0&0\\
			9&6&-1
		\end{smatrix}$.\par
		The characteristic polynomial is
		\[
			p(t)=\det(B-\lambda I)=-t
			\begin{vmatrix}
				5-t&-1\\
				9&-1-t
			\end{vmatrix}
			=-t(t^2-5t+t-5+9)
			=-(0-t)^{\textcolor{blue}{1}}(2-t)^{\textcolor{red}{2}}
		\]
		\begin{description}
		  \item[\normalfont$\mult 0=\textcolor{blue}{1}$:] Indeed $K_0=\cN(B-0I)^{\textcolor{blue}{1}}
		  =\cN(B)=
		  \Span\sthreevec 1{-1}3$ is the eigenspace $E_0$.
		  \item[\normalfont$\mult 2=\textcolor{red}{2}$:] This time, the generalized eigenspace is
		  \[
		  	K_2=\cN(B-2I)^{\textcolor{red}{2}} 
		  	=\cN
		  	\begin{pmatrix}
		  		3&2&-1\\
		  		0&-2&0\\
		  		9&6&-3
		  	\end{pmatrix}
		  	^2 =\cN
		  	\begin{pmatrix}
		  		0&-4&0\\
		  		0&4&0\\
		  		0&-12&0
		  	\end{pmatrix}
		  	=\Span\left\{\threevec 100,\threevec 001\right\}
		  \]
		  The corresponding eigenspace $\smash[b]{E_2=\Span\sthreevec 103}$ is a \emph{one}-dimensional subspace of $K_2$. The matrix is non-diagonalizable.
		\end{description}
		Observe also that $\R^3=\textcolor{blue}{K_0}\oplus \textcolor{red}{K_2}$.
	\end{enumerate}
\end{examples}


\pagebreak[3]


\boldsubsubsection{Properties of Generalized Eigenspaces and the Proof of Theorem \ref{thm:generalized1}}

A lot of work is required to justify the main result. Feel free to skip the proofs at first reading.

\begin{lemm}{}{gevectorbasic}
	Let $\lambda$ be an eigenvalue of $\rT\in\cL(V)$. Then:
	\begin{enumerate}
	  \item $E_\lambda$ is a subspace of $K_\lambda$, which is itself a subspace of $V$.
	  \item $K_\lambda$ is $\rT$-invariant.
	  \item Suppose $K_\lambda$ is finite-dimensional and $\mu\neq \lambda$. Then:
	  \begin{enumerate}
	    \item $K_\lambda$ is $(\rT-\mu \I)$-invariant and the restriction of $\rT-\mu\I$ to $K_\lambda$ is an isomorphism.
	    \item If $\mu$ is another eigenvalue, then $K_\lambda\cap K_\mu=\{\V0\}$. In particular $K_\lambda$ contains no eigenvectors other than those in $E_\lambda$. 
		\end{enumerate}
	\end{enumerate}
\end{lemm}

\begin{proof}
	Some of the details are left to Exercise \ref{exs:genevecbasic}.
	\begin{enumerate}
	  \item These are an exercise.
	  
	  \item Let $\vx\in K_\lambda$. We must show that $\rT(\vx)\in K_\lambda$. We know $\exists k$ such that $(\rT-\lambda \I)^k(\vx)=\V0$, but then
	  \begin{align*}
		  (\rT-\lambda \I)^k\bigl(\rT(\vx)\bigr)&=(\rT-\lambda \I)^{k}\bigl(\rT(\vx)-\lambda \vx+\lambda\vx\bigr)\\
		  &=(\rT-\lambda \I)^{k+1}(\vx)+\lambda(\rT-\lambda \I)^k(\vx) =\V0
	  \end{align*}
	  Otherwise said, $\rT(\vx)\in K_\lambda$.
	  
	  \item\begin{enumerate}
	    \item Since $K_\lambda$ is $\rT$-invariant (part 2), it is also $(\rT-\mu \I)$-invariant: if $\vx\in K_\lambda$, then
	    \[
	    	(\rT-\mu\I)(\vx)=\rT(\vx)-\mu\vx\in K_\lambda
	    \]
	    Suppose, for contradiction, that $\rT-\mu\I$ is \emph{not injective} on $K_\lambda$. Then
			\[
				\exists\vy\in K_\lambda\setminus\{\V0\}
				\text{ such that }
				(\rT-\mu \I)(\vy)=\V0
			\]
			Let $k\in\N$ be \textbf{minimal} such that $(\rT-\lambda \I)^k(\vy)=\V0$ ($\vy\in K_\lambda$!) and define $\vz=(\rT-\lambda \I)^{k-1}(\vy)\neq\V0$. We compute both $(\rT-\lambda\I)(\vz)$ and $(\rT-\mu\I)(\vz)$:
			\begin{itemize}
			  \item $(\rT-\lambda\I)(\vz)=(\rT-\lambda \I)^k(\vy)=\V0$ says $\vz$ is an eigenvector with eigenvalue $\lambda$.
				\item Since $\rT-\mu\I$ and $\rT-\lambda\I$ commute (easy exercise),
				\begin{align*}
					(\rT-\mu \I)(\vz)
					&=(\rT-\mu\I)(\rT-\lambda \I)^{k-1}(\vy)\\
					&=(\rT-\lambda \I)^{k-1}(\rT-\mu \I)(\vx)
					=\V0
				\end{align*}
				whence $\vz$ is also an eigenvector with eigenvalue $\mu$! If $\mu$ isn't an eigenvalue, then we already have a contradiction. Even if it is, $E_\mu\cap E_\lambda=\{\V0\}$ contradicts $\vz\neq\V0$.
			\end{itemize}
			\smallbreak
			We conclude that the restriction $(\rT-\mu\I)_{K_\lambda}\in\cL(K_\lambda)$ is injective. Since $\dim K_\lambda<\infty$, the restriction is automatically an isomorphism.
			\item	This is another exercise.\qedhere
	  \end{enumerate} 
	\end{enumerate}
\end{proof}


\pagebreak[3]


\begin{proof}[Proof of Theorem \ref{thm:generalized1}]
	Assume that the characteristic polynomial splits: $p(t)=(\lambda_1-t)^{m_1}\cdots (\lambda_k-t)^{m_k}$.
	\begin{description}
		\item[\normalfont{1. (a)}] Fix an eigenvalue $\lambda$. By definition, $\cN(\rT-\lambda \I)^m\le K_\lambda =\bigcup_{n\in\N}\cN(\rT-\lambda \I)^n$.\par
		The opposite inclusion follows from Lemma \ref{lemm:gevectorbasic}). Since $K_\lambda$ contains no eigenvectors from any eigenvalue other than $\lambda$, the characteristic polynomial of the restriction $\at\rT{K_\lambda}$ must be
		\[
			p_{K_\lambda}(t)=(\lambda-t)^{\dim K_\lambda}
		\]
		By $\rT$-invariance, $p_{K_\lambda}(t)$ divides $p(t)$ (Theorem \ref{thm:invariantsub}), whence $\dim K_{\lambda}\le m$.	By Cayley--Hamilton (Theorem \ref{thm:cayley-hamilton}), $p_{K_\lambda}(\at\rT{K_\lambda})=0$, whence
		\[
			\forall \vx\in K_\lambda,\  \left(\lambda \I-\rT\right)^{\dim K_\lambda}(\vx)=\V0 \implies K_\lambda\le \cN(\rT-\lambda \I)^m
		\]
		
		\item[\normalfont{1. (b) and 2.}] We prove simultaneously by induction\footnotemark{} on the number of distinct eigenvalues of $\rT$.
		\begin{description}
			\item[\normalfont\emph{Base case}] If $\rT$ has only one eigenvalue, then $p(t)=(\lambda-t)^m$ where $m=\dim V$. Cayley--Hamilton says that $(\rT-\lambda \I)^m(\vx)=\V0$ for all $\vx\in V$. Thus $V=K_\lambda$ and $\dim K_\lambda=m$.		
			\item[\normalfont\emph{Induction step}] Fix $k$ and suppose the results hold for all maps with $k$ distinct eigenvalues. Let $\rT$ have distinct eigenvalues $\lambda_1,\ldots,\lambda_k,\mu$, with multiplicities $m_1,\ldots,m_k,m$ respectively. Define
			\[
				W:=\cR(\rT-\mu\I)^m
			\]
			The subspace $W$ has several properties, the first two of which are left to Exercise \ref{exs:genevecproof}:
			\begin{itemize}
			  \item $W$ is $\rT$-invariant.
			  \item $W\cap K_\mu=\{\V0\}$ so that $\mu$ is not an eigenvalue of the restriction $\at\rT W$.
			  \item Each $K_{\lambda_j}\le W$: since $(\rT-\mu\I)_{K_{\lambda_j}}$ is an isomorphism (Lemma part 3), we can invert,
			  \[
			  	\smash{\vx\in K_{\lambda_j}
			  	\implies \vx=(\rT-\mu\I)^{m}\Bigl((\rT-\mu\I)_{K_{\lambda_j}}^{-1}\Bigr)^m(\vx) 
			  	\in \cR(\rT-\mu\I)^{m}=W}
			  \]
			\end{itemize}
			We conclude that each $\lambda_j$ is an eigenvalue of the restriction $\at\rT W$ with generalized eigenspace $K_{\lambda_j}$. Since $\at\rT W$ has $k$ distinct eigenvalues, the induction hypotheses apply:
			\[
				\smash{W=K_{\lambda_1}\oplus\cdots\oplus K_{\lambda_{k}}\quad\text{and}\quad p_{W}(t)=(\lambda_1-t)^{\dim K_{\lambda_1}}\cdots (\lambda_{k}-t)^{\dim K_{\lambda_{k}}}}
			\]
			Since $W\cap K_\mu=\{\V0\}$, we finish by using the rank--nullity theorem to count dimensions:\vspace{-8pt}
			\begin{align*}
				\dim V&=\Rank (\rT-\mu\I)^{m}+\Null (\rT-\mu\I)^{m}=\dim W+\dim K_\mu =\smash[b]{\sum_{j=1}^k}\dim K_{\lambda_j}+\dim K_\mu\\
				&\le m_1+\cdots+m_k+m =\operatorname{deg}(p(t))=\dim V
			\end{align*}
			We therefore have \emph{equality}; each $\dim K_{\lambda_j}=m_j$ and $\dim K_\mu=m$. We conclude that
			\[
				V=K_{\lambda_1}\oplus\cdots\oplus K_{\lambda_{k}}\oplus K_\mu
			\]
			which completes the induction step and thus the proof. Whew!\qedhere
		\end{description}
	\end{description}
\end{proof}

\footnotetext{%
	This is yet another argument where we consider a suitable subspace to which we can apply an induction hypothesis; recall the spectral theorem, Schur's lemma, bilinear form diagonalization, etc. Theorem \ref{thm:jordanbasis} will provide one more!%
}


\pagebreak[3]


\boldsubsubsection{Cycles of Generalized Eigenvectors}

Theorem \ref{thm:generalized1} says that every linear map whose characteristic polynomial splits has a generalized eigenbasis. This isn't quite the same as a Jordan canonical basis, but we're very close!

\begin{example}{}{}
	The matrix $A=
	\begin{smatrix}
		5&1&0\\
		0&5&1\\
		0&0&5
	\end{smatrix}
	\in M_3(\R)$ is a single Jordan block. There is a single generalized eigenspace $K_5=\R^3$, and the standard basis $\epsilon=\{\ve_1,\ve_2,\ve_3\}$ is Jordan canonical.\smallbreak
	The crucial observation for what follows is that one of these vectors $\ve_3$ \emph{generates} the others via repeated applications of $A-5I$:
	\[
		\ve_2=(A-5I)\ve_3,\qquad \ve_1=(A-5I)\ve_2=(A-5I)^2\ve_3
	\]
\end{example}

In the arguments and examples that follow, it will be helpful to refer to the linear map $\rU=\rT-\lambda\I$, where $\lambda$ is an eigenvalue of $\rT$. Note the following:
\begin{itemize}
  \item The nullspace of $\rU$ is the \emph{eigenspace}: $\cN(\rU)=E_\lambda\le K_\lambda$.
  \item $\rT$ and $\rU$ commute, that is $\rT\rU=\rU\rT$.
\end{itemize}

\begin{defn}{}{}
	Let $\rT$ have eigenvalue $\lambda$, suppose $\vx\in K_\lambda$ is a generalized eigenvector and let $k$ be minimal such that $(\rT-\lambda\I)^k(\vx)=\V0$. The \emph{cycle of generalized eigenvectors generated by $\vx$} is the set
	\[
		\beta_\vx:=\left\{(\rT-\lambda \I)^{k-1}(\vx),\ldots,(\rT-\lambda \I)(\vx),\vx\right\}
		=\left\{\rU^{k-1}(\vx),\ldots,\rU(\vx),\vx\right\}
	\]
\end{defn}

Note that only the first element $(\rT-\lambda\I)^{k-1}(\vx)$ of $\beta_{\vx}$ is an \emph{eigenvector}. Moreover,  $\Span\beta_\vx=\ip{\vx}$ is the $\rU$-cyclic subspace generated by $\vx$.\smallbreak

Our goal is to show that $K_\lambda$ has a basis consisting of cycles of generalized eigenvectors; putting these together results in a Jordan canonical basis.

\begin{lemm}{}{gencyclesbasic}
	Let $\beta_\vx$ be a cycle of generalized eigenvectors of $\rT$ with length $k$. Then:
	\begin{enumerate}
	  \item $\beta_\vx$ is linearly independent and thus a basis of $\Span\beta_\vx$ (which necessarily has dimension $k$).
	  \item $\Span\beta_\vx$ is $\rT$-invariant. With respect to $\beta_\vx$, the matrix of the restriction of $\rT$ is the $k\times k$ Jordan block $[\rT_{\Span\beta_\vx}]_{\beta_\vx}=
	  \begin{smatrix}
		  \lambda&1&&\\
		  &\lambda&\ddots&\\
		  &&\ddots&1\\
		  &&&\lambda
	  \end{smatrix}$.
	\end{enumerate}
\end{lemm}



\begin{proof}
	As indicated above, we write $\rU=\rT-\lambda\I$.
	\begin{enumerate}
	  \item Feed the linear combination $\sum_{j=0}^{k-1}a_j\rU^j(\vx)=\V0$ to $\rU^{k-1}$ to obtain
	  \[
	  	a_0\rU^{k-1}(\vx)=\V0\implies a_0=0
	  	\tag{remember $\rU^k(\vx)=\V0$!}
	  \]
	  Now feed the same linear combination to $\rU^{k-2}$, etc., to see that all coefficients $a_j=0$.
	  \item Since $\rT$ and $\rU$ commute,
	  \[
	  	\rT\bigl(\rU^j(\vx)\bigr)=\rU^j\bigl(\rT(\vx)\bigr)=\rU^j\bigl((\rU+\lambda\I)(\vx)\bigr)=\rU^{j+1}(\vx)+\lambda\rU^j(\vx)\in\Span\beta_\vx
	  \]
	  This justifies both $\rT$-invariance and the Jordan block claim!\qedhere
	\end{enumerate}
\end{proof}


\pagebreak[3]


Once we have the generalized eigenspaces of $\rT$, the naïve approach to finding a Jordan canonical basis is to play with cycles until we find a basis for each $K_\lambda$. \emph{Many} choices of canonical basis exist for a given map! We'll consider a more systematic method in the next section.

\begin{examples}{}{jordansimple1}
	We compute Jordan canonical bases for three linear maps/matrices.
	\begin{enumerate}
	  \item\label{ex:jordansimple11} The characteristic polynomial $p(t)=(1-t)^3$ of $A=
	  \begin{smatrix}
			1&0&2\\
			0&1&6\\
			6&-2&1
		\end{smatrix}$
		splits over $\R$.\par
		We hack our way to a Jordan canonical basis by choosing some $\vx\in K_1=\R^3$ and considering what $U=A-I$ does to it! For instance, with $\vx=\ve_1$,
		\[
			\beta_\vx
			=\left\{\rU^2\threevec 100,\rU\threevec 100,\threevec 100\right\}
			=\left\{\threevec{12}{36}0,\threevec 006,\threevec 100\right\}
		\]
		is a Jordan canonical basis. Otherwise said,
		\[
			A=QJQ^{-1}=
			\begin{pmatrix}
				12&0&1\\
				36&0&0\\
				0&6&0
			\end{pmatrix}
			\begin{pmatrix}
				1&1&0\\
				0&1&1\\
				0&0&1
			\end{pmatrix}
			\begin{pmatrix}
				12&0&1\\
				36&0&0\\
				0&6&0
			\end{pmatrix}
			^{-1}
		\] 
		In practice, almost any choice of $\vx\in\R^3$ will generate a cycle of length three and thus a basis!
		
	  \item The matrix $B=
	  \begin{smatrix}
			7&1&-4\\
			0&3&0\\
			8&1&-5
		\end{smatrix}
		\in M_3(\R)$ has characteristic polynomial $p(t)=-(t+1)^{\textcolor{blue}1}(t-3)^{\textcolor{red}2}$.\smallbreak
		Since $\dim K_{-1}=\textcolor{blue}1$, we see that $K_{-1}=E_{-1}=\smash{\Span\sthreevec 102}$ is spanned by a cycle of length one.\smallbreak
	Since $\dim K_3=\textcolor{red}2$, we have
	  \[
	  	K_3=\cN(B-3I)^{\textcolor{red}2}
	  	=\cN
	  	\begin{pmatrix}
	  		4&1&-4\\
	  		0&0&0\\
	  		8&1&-8
	  	\end{pmatrix}
	  	^2=
	  	\cN
	  	\begin{pmatrix}
	  		-16&0&16\\
	  		0&0&0\\
	  		-32&0&32
	  	\end{pmatrix}
	  	=\Span\left\{\threevec 101,\threevec 010\right\}
	  \]
	  which is spanned by a cycle of length two: $\sthreevec 101=(B-3I)\sthreevec 010$ is an eigenvector.\smallbreak
	  Putting it together, $\beta=\left\{\sthreevec 102,\sthreevec 101,\sthreevec 010\right\}$ is a Jordan canonical basis for $B$, and
		\[
			B=QJQ^{-1}=
			\begin{pmatrix}
				1&1&0\\
				0&0&1\\
				2&1&0
			\end{pmatrix}
			\begin{pmatrix}
				-1&0&0\\
				0&3&1\\
				0&0&3
			\end{pmatrix}
			\begin{pmatrix}
				1&1&0\\
				0&0&1\\
				2&1&0
			\end{pmatrix}
			^{-1}
		\]
	 
		\item Let $\rT=\diff x\in\cL\big(P_3(\R)\big)$. With respect to the standard basis $\epsilon=\{1,x,x^2,x^3\}$, $[\rT]_\epsilon=
		\begin{smatrix}
			0&1&0&0\\
			0&0&2&0\\
			0&0&0&3\\
			0&0&0&0
		\end{smatrix}$.\par
		We have a single generalized eigenspace $K_0=P_3(\R)$. It is easy to check that $f(x)=x^3$ generates a cycle of length three and thus a Jordan canonical basis:
		\[
			\beta=\big\{6,6x,3x^2,x^3\big\}
			\implies[\rT]_\epsilon =
% 			\begin{smatrix}				
% 				0&1&0&0\\
% 				0&0&2&0\\
% 				0&0&0&3\\
% 				0&0&0&0
% 			\end{smatrix}
% 			=
			\begin{pmatrix}
				6&0&0&0\\
				0&6&0&0\\
				0&0&3&0\\
				0&0&0&1
			\end{pmatrix}
			\begin{pmatrix}
				0&1&0&0\\
				0&0&1&0\\
				0&0&0&1\\
				0&0&0&0
			\end{pmatrix}
			\begin{pmatrix}
				6&0&0&0\\
				0&6&0&0\\
				0&0&3&0\\
				0&0&0&1
			\end{pmatrix}
			^{-1}
		\]
	\end{enumerate}
\end{examples}


\pagebreak[3]


Our final results state that this process works in general.

\begin{thm}{}{jordanbasis}
	Let $\rT\in\cL(V)$ have an eigenvalue $\lambda$. If $\dim K_\lambda<\infty$, then there exists a basis $\beta_\lambda=\beta_{\vx_1}\cup\cdots\cup\beta_{\vx_n}$ of $K_\lambda$ consisting of finitely many linearly independent cycles.
\end{thm}

Intuition suggests that we start with a basis of the eigenspace $E_\lambda$ and extend backwards: for each $\vx$, if $\vx=(\rT-\lambda\I)(\vy)$, then $\vx\in\beta_\vy$; now repeat until you have a maximum length cycle. This is essentially what we do, though some sneakiness is required to keep track of everything.

\begin{proof}
	We prove by induction on $m=\dim K_\lambda$.
	\begin{description}
		\item[\normalfont\emph{Base case} ($m=1$)] $K_\lambda=E_\lambda=\Span\vx$ for some eigenvector $\vx$. Plainly $\{\vx\}=\beta_\vx$ does the job.
		\item[\normalfont\emph{Induction step}] Fix $m\ge 2$, and suppose \emph{every} generalized eigenspace with dimension $<m$ (for \emph{any} linear map!) has a basis consisting of independent cycles of generalized eigenvectors.\par
		Write $n=\dim E_\lambda\le m$ and $\rU=(\rT-\lambda\I)_{K_\lambda}$. We break the argument into several steps. 
	\end{description}
	\begin{enumerate}
	  \item Define $W=E_\lambda\cap \cR(\rU)$: that is
	  \[
	  	\vw\in W\iff
	  	\rU(\vw)=\V0\ \text{ and }\ 
	  	\vw=\rU(\vv)\text{ for some }\vv\in K_\lambda
	  \]
	  Let $k=\dim W$, choose a subspace $X$ such that $E_\lambda=W\oplus X$ and select a basis $\{\vx_{k+1},\ldots,\vx_n\}$ of $X$.
	   If $k=0$, the induction step is finished (Exercise \ref{exs:jordanproof} shows that $K_\lambda=E_\lambda$\ldots).
	  
	  
	  \item Otherwise, take $j=1$ in the proof of Lemma \ref{lemm:gencyclesbasic} to see that $\cR(\rU)$ is $\rT$-invariant; it follows that $\cR(U)$ is the single generalized eigenspace (eigenvalue $\lambda$) of the restriction $\rT_{\cR(\rU)}$. By the rank--nullity theorem,
		\[
			\dim \cR(\rU)=\Rank\rU=\dim K_\lambda-\Null\rU=m-\dim E_\lambda <m
		\]
		By the induction hypothesis, $\cR(\rU)$ has a basis of independent cycles. Since the last non-zero element in each cycle is an eigenvector, this basis consists of $k$ distinct cycles $\beta_{\hat\vx_1}\cup\cdots\cup\beta_{\hat\vx_k}$ whose terminal vectors form a basis of $W$.
		
	  \item Since each $\hat\vx_j\in\cR(\rU)$, there exist vectors $\vect xk$ such that $\hat\vx_j=\rU(\vx_j)$. Including the length-one cycles generated by the basis of $X$, the cycles $\beta_{\vx_1},\ldots,\beta_{\vx_n}$ now contain
	  \[
	  	\dim\cR(\rU)+ \underbrace{\quad k\quad}_{\vect xk} +\underbrace{(n-k)}_{\vx_{k+1},\ldots,\vx_n}
	  	=\Rank\rU+\Null\rU=m
	  \]
	  vectors. Exercise \ref{exs:jordanproof} verifies that these vectors are linearly independent.\qedhere
	\end{enumerate} 
\end{proof}


By Theorem \ref{thm:generalized1}, if the characteristic polynomial of $\rT\in\cL(V)$ splits, then $V=K_{\lambda_1}\oplus\cdots\oplus K_{\lambda_k}$ is the direct sum of generalized eigenspaces. Since each $K_{\lambda_j}$ now has a basis $\beta_\lambda$ consisting of finitely many cycles, Lemma \ref{lemm:gencyclesbasic} says that the matrix $[\rT_{K_\lambda}]_{\beta_\lambda}$ is a Jordan canonical form, whence $\beta=\bigcup\beta_\lambda$ is a Jordan canonical basis for $\rT$. In conclusion:

\begin{cor}{}{jordanbasiscor}
	If the characteristic polynomial of $\rT$ splits, then it has a Jordan canonical basis.
\end{cor}


\pagebreak[3]


\begin{exercises}
	\hangindent\leftmargini\emph{Key concepts:\quad Jordan blocks, canonical forms \& bases,\quad Generalized eigenvectors \& spaces,\quad Building Jordan bases using cycles of generalized eigenvectors}
	
	
	\begin{enumerate}
	  \item Compute the generalized eigenspaces $K_\lambda$, find bases consisting of unions of disjoint cycles of generalized eigenvectors, and thus find a Jordan canonical form $J$ and invertible $Q$ so as to express the matrix in the form $QJQ^{-1}$.
		\begin{enumerate}
	    \item $A=
	    \begin{smatrix}
	    	1&1\\
	    	-1&3
	    \end{smatrix}$
	    \qquad
	    (b)\ \ $B=\begin{smatrix}
	    	1&2\\
	    	3&2
	    \end{smatrix}$
			\qquad
	    (c)\ \ $C=\begin{smatrix}
	    	11&-4&-5\\
	    	21&-8&-11\\
	    	3&-1&0
	    \end{smatrix}$
	    \qquad
	    (d)\ \ $D=\begin{smatrix}
	    	2&1&0&0\\
	   		0&2&1&0\\
	    	0&0&3&0\\
	    	0&1&-1&3
	    \end{smatrix}$
		\end{enumerate}
		
		
		\item If $\beta=\{\vect vn\}$ is a Jordan canonical basis, what can you say about $\vv_1$?
		
		
		\item Briefly explain why matrix $A=
		\begin{smatrix}
			0&-1\\
			1&0
		\end{smatrix}$
		has no Jordan canonical form $J\in M_2(\R)$.
		
		
		\item Find a Jordan canonical basis for each linear map $\rT$:
		\begin{enumerate}
		  \item $\rT\in\cL\big(P_2(\R)\big)$ defined by $\rT\big(f(x)\big)=2f(x)-f'(x)$
		  \item $\rT(f)=f'$ defined on $\Span\{1,t,t^2,e^t,te^t\}$
		  \item $\rT(A)=2A+A^T$ defined on $M_2(\R)$
		\end{enumerate}
		
		
		\item In Example \ref*{ex:jordansimple1}.\ref*{ex:jordansimple11}, suppose $\vx=\sthreevec abc$. Show that almost any choice of $a,b,c$ produces a Jordan canonical basis $\beta_\vx$.
	
	  
	  \item\label{exs:genevecbasic} Complete the proof of Lemma \ref{lemm:gevectorbasic}.
	  \begin{enumerate}
	    \item Prove part 1, that $E_\lambda\le K_\lambda\le V$.
	    \item Verify that $\rT-\mu\I$ and $\rT-\lambda\I$ commute.
	    \item Prove part 3(b): generalized eigenspaces for distinct eigenvalues have trivial intersection.
	  \end{enumerate}
	  
	  
	  \item\label{exs:genevecproof} Consider the induction step in the proof of Theorem \ref{thm:generalized1}.
	  \begin{enumerate}
	    \item Prove that $W$ is $\rT$-invariant.
	    \item Explain why $W\cap K_\mu=\{\V0\}$.
	    \item The assumption $p_W(t)=(\lambda_1-t)^{\dim K_{\lambda_1}}\cdots (\lambda_k-t)^{\dim K_{\lambda_k}}$ near the end of the proof is the induction hypothesis for part 1(b). Why can't we also assume that $\dim K_{\lambda_j}=m_j$ and thus tidy the inequality argument near the end of the proof?
	  \end{enumerate}
	  
	  
	  \item\label{exs:jordanproof} We finish some of the details of Theorem \ref{thm:jordanbasis}.
	  \begin{enumerate}
	    \item In step 1, suppose $\dim W=k=0$. Explain why $\{\vx_1,\ldots,\vx_n\}$ is in fact a basis of $K_\lambda$, so that the rest of the proof is unnecessary.
	    \item In step 3, prove that the $m$ vectors in the cycles $\beta_{\vx_1},\ldots,\beta_{\vx_n}$ are linearly independent.\par
	    (\emph{Hint: model your argument on part 1 of Lemma \ref{lemm:gencyclesbasic}})
	  \end{enumerate}
	 
	\end{enumerate}
\end{exercises}



\clearpage



\subsection{Cycle Patterns and the Dot Diagram}


In this section we develop a method for computing Jordan forms more efficiently and systematically. We start with a lengthy example that provides some clues how to proceed.

\begin{example}{}{dotdiagmotiv}
	Precisely three Jordan canonical forms $A,B,C\in M_3(\R)$ correspond to the characteristic polynomial $p(t)=(5-t)^3$:
	\[
		A=
		\begin{pmatrix}
			5&0&0\\
			0&5&0\\
			0&0&5
		\end{pmatrix}
		\qquad
		B=\begin{pmatrix}
			5&1&0\\
			0&5&0\\
			0&0&5
		\end{pmatrix}
		\qquad
		C=\begin{pmatrix}
			5&1&0\\
			0&5&1\\
			0&0&5
		\end{pmatrix}
	\]
	In all three cases the standard basis $\beta=\{\ve_1,\ve_2,\ve_3\}$ is Jordan canonical, so how do we distinguish things? By considering the \emph{\textcolor{blue}{number}} and \emph{\textcolor{red}{lengths}} of the cycles of generalized eigenvectors.
	\begin{itemize}
	  \item $A$ has eigenspace $E_5=K_5=\R^3$. Since $(A-5I)\vv=\V0$ for all $\vv\in\R^3$, we have \textcolor{red}{maximum cycle-length one}. We therefore need \textcolor{blue}{three distinct cycles} to construct a Jordan basis, e.g.
	  \[
		  \beta_{\ve_1}=\{\ve_1\},\quad 
		  \beta_{\ve_2}=\{\ve_2\},\quad 
		  \beta_{\ve_3}=\{\ve_3\} 
		  \implies \beta=\beta_{\ve_1}\cup\beta_{\ve_2}\cup\beta_{\ve_3}
		  =\{\ve_1,\ve_2,\ve_3\}
	  \]
	  
	  \item $B$ has eigenspace $E_5=\Span\{\ve_1,\ve_3\}$. Provided $\vv\not\in E_5$,
	  \[
	  	\vv=\threevec abc\implies (B-5I)\vv=\threevec{b}00 \implies (B-5I)^2\vv=\V0
	  \]
	  shows that $\beta_\vv$ is a cycle with \textcolor{red}{maximum length two}. We therefore need \textcolor{blue}{two distinct cycles} with lengths \textcolor{red}{two} and \textcolor{red}{one} to construct a Jordan basis, e.g.
	  \[
	  	\beta_{\ve_2}=\bigl\{(B-5I)\ve_2,\,\ve_2\bigr\}=\{\ve_1,\ve_2\},\quad 
	  	\beta_{\ve_3}=\{\ve_3\}
	  	\implies \beta=\beta_{\ve_2}\cup\beta_{\ve_3}=\{\ve_1,\ve_2,\ve_3\}
	 	\]
	  
	  \item $C$ has eigenspace $E_5=\Span\ve_1$. This time
	  \[
	  	\vv=\threevec abc
	  	\implies (C-5I)\vv=\threevec{b}c0,\quad 
	  	(C-5I)^2\vv=\threevec c00,\quad (C-5I)^3\vv=\V0
	  \]
	  Provided $c\neq 0$, this generates a cycle with \textcolor{red}{maximum length three}. Indeed this cycle is a Jordan basis, so \textcolor{blue}{one cycle} is all we need:
	  \[
	  	\beta=\beta_{\ve_3}=\bigl\{(C-5I)^2\ve_3,\,(C-5I)\ve_3,\,\ve_3\bigr\}=\{\ve_1,\ve_2,\ve_3\}
	  \]\vspace{-5pt}
	\end{itemize}
\end{example}

Why is the example relevant? If $\rT\in\cL(V)$ has characteristic polynomial $p(t)=(5-t)^3$, Theorem \ref{thm:jordanbasis} tells us that $\rT$ has a Jordan canonical form: one of the matrices $A,B,C$. Our goal is to develop a method whereby the pattern of cycle-lengths can be determined, thus allowing us to discern which Jordan form is correct. As a side-effect, this will also demonstrate that the pattern of cycle lengths for a given $\rT$ is independent of the Jordan basis so that (modulo some reasonable restriction) the Jordan \emph{form} of $\rT$ is unique. To aid us in this endeavor, we require some terminology\ldots


\pagebreak[3]


\begin{defn}{}{dotdiag}
	Let $K_\lambda$ be a generalized eigenspace of $\rT\in\cL(V)$ and suppose  $\beta_\lambda=\beta_{\vx_1}\cup\cdots\cup\beta_{\vx_n}$ is a Jordan canonical basis of $\at\rT{K_\lambda}$, whose cycles are arranged in \emph{non-increasing} length. That is:
	\begin{enumerate}
	  \item $\beta_{\vx_j}=\big\{(\rT-\lambda \I)^{k_j-1}(\vx_j),\ldots,\vx_j\big\}$ has length $k_j$, \ and,
	  \item $k_1\ge k_2\ge\cdots\ge k_n$
	\end{enumerate}
	The \emph{dot diagram} of $\at\rT{K_\lambda}$ is a representation of the elements of $\beta_\lambda$, one dot for each vector: the $j\th$ column represents the elements of $\beta_{\vx_j}$ arranged vertically with $\vx_j$ at the bottom.
\end{defn}

Given a linear map, our goal is to identify the dot diagram as an intermediate step in the computation of a Jordan basis. First we observe how the conversion of dot diagrams to Jordan forms is essentially trivial.

\begin{example}{}{}
	Suppose $\rT\in\cL(\R^{12})$ has the following eigenvalues and dot diagrams:
  \[
	  \begin{array}{c@{\qquad}c@{\qquad}c}
	  	\lambda_1=-4&\lambda_2=7&\lambda_3=9\\
	  	\begin{array}{cc}
	  		\bullet&\bullet\\
	  		\bullet&\bullet\\
	  		\bullet
	  	\end{array}
	  	&
	  	\begin{array}{cc}
	  		\bullet&\bullet\\
	  		\bullet\\
	 			\bullet
	 		\end{array}
	 		&
	  	\begin{array}{ccc}
	  		\bullet&\bullet&\bullet\\
	  		&&\\
	  		&&
	  	\end{array}
	  \end{array}
	\]
	\begin{minipage}[t]{0.74\linewidth}\vspace{0pt}
  	Consider the diagram for $\lambda_1=-4$. Each dot depicts a single vector in basis $\beta_{-4}=\beta_{\vx_1}\cup\beta_{\vx_2}$ of $K_{-4}$, as shown on the right.
  	\begin{enumeratea}\itemsep0pt
  	  \item The columns are the cycles: $\dim K_{-4}=3+2=5$.
  	  \item The top row is the eigenspace: $E_{-4}=\Span\big\{(\rT+4\I)^2\vx_1,(\rT+4\I)\vx_2\big\}$ has dimension 2. 
  	  \item The maximum cycle length is three: $K_{-4}=\cN(\rT+4\I)^3$.
  	\end{enumeratea}
	\end{minipage}
	\hfill
	\begin{minipage}[t]{0.25\linewidth}\vspace{0pt}
	  \hfill$\begin{array}[t]{@{}cc@{}}
			(\rT+4\I)^2\vx_1&(\rT+4\I)\vx_2\\
			(\rT+4\I)\vx_1&\vx_2\\
			\vx_1
		\end{array}$
	\end{minipage}\bigbreak
	Similarly for $\lambda_2=7$: $\dim K_7=4$, $\dim E_7=2$ and $K_7=\cN(\rT-7\I)^3$.\par
	Finally, for $\lambda_3=9$ the generalized eigenspace is the eigenspace: $K_9=E_9=\cN(\rT-9\I)$ has dimension 3.\smallbreak
	The map $\rT$ has a Jordan canonical basis $\beta$ with respect to which its Jordan canonical form is
	\[
		[\rT]_\beta =\left(
		\begin{array}{@{}c@{}c@{}c@{}c@{}c@{}c@{}c@{}c@{}}
			\begin{array}{@{}ccc|@{}}
				-4 & 1 & 0\\
 				0 & -4 & 1\\
 				0 & 0 & -4\\\hline
			\end{array}
			\\
 			&
 			\begin{array}{@{}|cc|@{}}
 				\hline
 				-4 & 1\\
 				0 & -4\\\hline
 			\end{array}
 			\\
			&&
			\begin{array}{|ccc|}\hline
 				7 & 1 & 0\\
 				0 & 7 & 1\\
 				0 & 0 & 7\\\hline
 			\end{array}
 			\\
 			&&&
 			\begin{array}{|c|}\hline
 				7\\\hline
 			\end{array}
 			\\
			&&&&
			\begin{array}{|c|}\hline
				9\\\hline
			\end{array}
			\\
			&&&&&
			\begin{array}{|c|}\hline
				9\\\hline
			\end{array}
			\\
			&&&&&&
			\begin{array}{|c}\hline
				9
			\end{array}
		\end{array}
		\right)
	\]
	Note how the sizes of the Jordan blocks are \emph{non-increasing} within each eigenvalue.
\end{example}


\pagebreak[3]


\begin{thm}{}{}
	Suppose $\beta_\lambda$ is a Jordan canonical basis of $\at\rT{K_\lambda}$ as described in Definition \ref{defn:dotdiag}, and suppose the $i$\th{} row of the dot diagram has $r_i$ entries. Then:
	\begin{enumerate}
	  \item For each $r\in\N$, the vectors associated to the dots in the first $r$ rows form a basis of $\cN(\rT-\lambda \I)^r$.
	  \item The dimension of the eigenspace is $r_1=\Null(\rT-\lambda\I)=\dim V-\Rank(\rT-\lambda \I)$.
	  \item When $i>1$, $r_i=\Null(\rT-\lambda \I)^{i}-\Null(\rT-\lambda \I)^{i-1}= \Rank(\rT-\lambda \I)^{i-1}-\Rank(\rT-\lambda \I)^{i}$
	\end{enumerate}
\end{thm}

\begin{example*}{\ref{ex:dotdiagmotiv} cont}{}
We describe the dot diagrams of the three matrices $A$, $B$, $C$, the corresponding vectors in the Jordan canonical basis $\beta$, and the values $r_i$.
\begin{itemize}
	\item[$A:$] $\begin{array}{ccc}
			\bullet&\bullet&\bullet
		\end{array}
		\qquad
		\begin{array}{ccc}
			\ve_1&\ve_2&\ve_3
		\end{array}$\par
		Since $A-5I$ is the zero matrix, $r_1=3-\Rank(A-5I)=3$. The dot diagram has one row, corresponding to three independent cycles of length one: $\beta=\beta_{\ve_1}\cup\beta_{\ve_2}\cup\beta_{\ve_3}$.
		
	\item[$B:$] $\begin{array}[t]{cc}
			\bullet&\bullet\\
			\bullet
		\end{array}
		\qquad
		\begin{array}[t]{cc}
			\ve_1&\ve_3\\
			\ve_2
		\end{array}
		\qquad
		\begin{array}[t]{cc}
			(B-5I)\ve_2&\ve_3\\
			\ve_2
		\end{array}$\smallbreak
		Row 1: \ $B-5I=
		\begin{smatrix}
			0&1&0\\
			0&0&0\\
			0&0&0
		\end{smatrix}$
		has rank 1, whence $r_1=3-1=2$. This row $\{\ve_1,\ve_3\}$ spans $E_5=\cN(B-5I)$.\par
		Row 2: \ $(B-5I)^2$ is the zero matrix, whence $r_2=\Rank(B-5I)-\Rank(B-5I)^2=1-0=1$.\smallbreak
		The dot diagram corresponds to $\beta=\beta_{\ve_2}\cup\beta_{\ve_3}=\{\ve_1,\ve_2\}\cup\{\ve_3\}$.
		
		
	\item[$C:$] $\begin{array}[t]{ccc}
		\bullet\\\bullet\\\bullet
		\end{array}
		\qquad
		\begin{array}[t]{c}
		\ve_1\\
		\ve_2\\
		\ve_3
		\end{array}
		\qquad
		\begin{array}[t]{c}
		(C-5I)^2\ve_3\\
		(C-5I)\ve_3\\
		\ve_3
		\end{array}$\smallbreak
		Row 1: \ $C-5I=
		\begin{smatrix}
			0&1&0\\
			0&0&1\\
			0&0&0
		\end{smatrix}$
		has rank 2, whence $r_1=3-2=1$. This row $\{\ve_1\}$ spans $E_5=\cN(C-5I)$.\par
		Row 2: \ $(C-5I)^2=
		\begin{smatrix}
			0&0&1\\
			0&0&0\\
			0&0&0
		\end{smatrix}$
		has rank 1, whence $r_2=2-1=1$. The first two rows $\{\ve_1,\ve_2\}$ form a basis of $\cN(C-5I)^2$.\par
		Row 3: \ $(C-5I)^3$ is the zero matrix, whence $r_3=\Rank(C-5I)^2-\Rank(C-5I)^3=1-0=1$.
	\end{itemize}
\end{example*}


\begin{proof}
As previously, let $\rU=\rT-\lambda\I$.
\begin{enumerate}
  \item Since each dot represents a basis vector $\rU^p(\vv_j)$, any $\vv\in K_\lambda$ may be written uniquely as a linear combination of the dots. Applying $\rU$ moves all dots up a row and all dots in the top row to $\V0$. It follows that $\vv\in \cN(\rU^r)\Longleftrightarrow$ it lies in the span of the first $r$ rows. Since the dots are linearly independent, they form a basis.
  \item By part 1, $r_1=\dim \cN(\rU)=\Null(\rT-\lambda \I)=\dim V-\Rank(\rT-\lambda \I)$.
  \item More generally,
  \begin{align*}
  r_i&=(r_1+\cdots+r_i)-(r_1+\cdots +r_{i-1})=\dim\cN(\rU^i)-\dim \cN(\rU^{i-1})\\
  &=\Null(\rU^i)-\Null(\rU^{i-1})=\Rank(\rT-\lambda \I)^{i-1}-\Rank(\rT-\lambda \I)^{i}\tag*{\qedhere}
  \end{align*}
\end{enumerate}
\end{proof}


\pagebreak[3]


Since the ranks of the maps $(\rT-\lambda\I)^i$ are independent of basis, so also is the dot diagram\ldots

\begin{cor}{}{jordanunique}
	For any eigenvalue $\lambda$, the dot diagram is uniquely determined by $\rT$ and $\lambda$. If Jordan blocks for each eigenspace are listed in non-increasing order, then the Jordan form of a linear map is unique up to the order of its eigenvalues.
\end{cor}

We now have a slightly more systematic method for finding Jordan canonical bases.

\begin{example}{}{jordanpenult}
	The matrix $A=
	\smash[b]{\begin{smatrix}
		6&2&-4&-6\\
		0&3&0&0\\
		0&0&3&0\\
		2&1&-2&-1
	\end{smatrix}}$ 
	has characteristic polynomial $p(t)=(2-t)^{\textcolor{blue}{1}}(3-t)^{\textcolor{red}{3}}$, and two generalized eigenspaces:
	\begin{description}
	  \item[$(\lambda=2)$] $K_2=\cN(A-2I)^{\textcolor{blue}{1}}$ is the one-dimensional eigenspace itself! The trivial dot diagram $\bullet$ corresponds to a single eigenvector in
	  \[
		  E_2=\cN
		  \begin{pmatrix}
				4&2&-4&-6\\
				0&1&0&0\\
				0&0&1&0\\
				2&1&-2&-3
			\end{pmatrix}
			=\Span\fourvec 3002
		\]
	  \item[$(\lambda=3)$] $K_3=\cN(A-3I)^3$. To find the dot diagram, compute powers of $A-3I$:\smallbreak
		Row 1: \ $A-3I=
		\begin{smatrix}
			3&2&-4&-6\\
			0&0&0&0\\
			0&0&0&0\\
			2&1&-2&-4
		\end{smatrix}$
		has rank 2 and the first row has $r_1=4-2=2$ entries.\par
		Row 2: \ $(A-3I)^2=
		\begin{smatrix}
			-3&0&0&6\\
			0&0&0&0\\
			0&0&0&0\\
			-2&0&0&4
		\end{smatrix}$
		has rank 1 and the second row has $r_2=2-1=1$ entry.\smallbreak
		Since we now have three dots ($=\dim K_3$), the algorithm terminates and we conclude that the dot diagram for $K_3$ is 
		$\begin{array}[t]{cc}
			\bullet&\bullet\\\bullet&
		\end{array}$\par
		For the single dot in the second row, choose anything in $\cN(A-3I)^2$ that isn't an eigenvector; perhaps the simplest choice is $\vx_1=\ve_2$, which yields the two-cycle
		\[
			\beta_{\vx_1}
			=\big\{(A-3I)\vx_1,\vx_1\big\} 
			=\left\{\sfourvec 2001,\sfourvec 0100\right\}
		\]
		To finish the first row, choose any eigenvector to complete the span: for instance $\vx_2=\smash[t]{\sfourvec 0210}$.
	\end{description}

	We now have suitable cycles and a Jordan canonical basis/form: 
	\[
		\beta=\left\{\fourvec 3002, \fourvec 2001, \fourvec 0100, \fourvec 0210\right\},
		\qquad
		A=QJQ^{-1}=
		\begin{pmatrix}
			3&2&0&0\\
			0&0&1&2\\
			0&0&0&1\\
			2&1&0&0
		\end{pmatrix}
		\begin{pmatrix}
			2&0&0&0\\
			0&3&1&0\\
			0&0&3&0\\
			0&0&0&3
		\end{pmatrix}
		\begin{pmatrix}
			3&2&0&0\\
			0&0&1&2\\
			0&0&0&1\\
			2&1&0&0
		\end{pmatrix}
		^{-1}
	\]
	Other choices of basis are available: for instance, we could generate a two-cycle $\beta_{\vx_1}$ using any vector $\vx_1=a\ve_2+b\ve_3+c(2\ve_1+\ve_4)\in\cN(A-2I)^2$ for which $a,b$ are not both zero\ldots
\end{example}

%\vfil

\pagebreak[3]



\begin{example}{}{}
	For variety, here is an example with a non-matrix map.\medbreak
	Let $\epsilon=\{1,x,y,x^2,y^2,xy\}$ and define the partial differential operator $\rT\bigl(f(x,y)\bigr)=2\partials[f]{x}-\partials[f]{y}$ on the real vector space $V=\Span_{\R}\epsilon$. We quickly compute:
	\[
		[\rT]_\epsilon=
		\begin{pmatrix}
			0 & 2 & -1 & 0 & 0 & 0\\
			0 & 0 & 0 & 4 & 0 & -1\\
			0 & 0 & 0 & 0 & -2 & 2\\
			0 & 0 & 0 & 0 & 0 & 0\\
			0 & 0 & 0 & 0 & 0 & 0\\
			0 & 0 & 0 & 0 & 0 & 0
		\end{pmatrix}
		\implies p(t)=t^6,\quad
		[\rT^2]_\epsilon=
		\begin{pmatrix}
			0 & 0 & 0 & 8 & 2 & -4\\
			0 & 0 & 0 & 0 & 0 & 0\\
			0 & 0 & 0 & 0 & 0 & 0\\
			0 & 0 & 0 & 0 & 0 & 0\\
			0 & 0 & 0 & 0 & 0 & 0\\
			0 & 0 & 0 & 0 & 0 & 0
		\end{pmatrix},
		\quad [\rT^3]_\epsilon=O
	\]
	There is only one eigenvalue $\lambda=0$ and therefore one generalized eigenspace $K_0=V$. We could keep working with matrices, but it is easy to translate their nullspaces back to subspaces of $V$, from which the necessary data can be read off:
	\begin{align*}
		&\cN(\rT)=\Span\big\{1,x+2y,x^2+4y^2+4xy\big\}\qquad 
		&&\Null \rT=3,\ \Rank \rT=3,\ r_1=3\\[2pt]
		&\cN(\rT^2)=\Span\big\{1,x,y,x^2+2xy,2y^2+xy\big\}\qquad 
		&&\Null \rT^2=5,\ \Rank \rT^2=1,\ r_2=3-1=2
	\end{align*}
	\begin{minipage}[t]{0.8\linewidth}\vspace{0pt}
		These calculations produce five dots; since $\dim K_0=6$, the last row has one dot and the dot diagram is as shown.
	\end{minipage}
	\hfill
	\begin{minipage}[t]{0.15\linewidth}\vspace{0pt}
		\hfill$\begin{array}[t]{ccc}
		\bullet&\bullet&\bullet\\
		\bullet&\bullet&\\
		\bullet&&
	\end{array}$
	\end{minipage}\smallbreak
	Since the first two rows span $\cN(\rT^2)$, we may choose any $f_1\not\in\cN(\rT^2)$ for the final dot. A suitable choice is $f_1=xy$, from which the first column of the dot diagram becomes
	\[
		\begin{matrix}
			\rT^2(xy)&\bullet&\bullet\\[2pt]
			\rT(xy)&\bullet&\\[2pt]
			xy&&
		\end{matrix}
		\qquad\quad
		\begin{matrix}
			-4&\bullet&\bullet\\[2pt]
			2y-x&\bullet&\\[2pt]
			xy&&
		\end{matrix}
	\]
	Now choose the second dot on the second row to be anything in $\cN(\rT^2)$ such that the first two rows span $\cN(\rT^2)$. This time $f_2=x^2-4y^2$ is suitable and the diagram becomes
	\[
		\begin{matrix}
			\rT^2(xy)&\rT(x^2-4y^2)&\bullet\\[2pt]
			\rT(xy)&x^2-4y^2&\\[2pt]
			xy&&
		\end{matrix}
		\qquad\quad
		\begin{matrix}
			-4&4x+8y&\bullet\\[2pt]
			2y-x&x^2-4y^2&\\[2pt]
			xy&&
		\end{matrix}
	\]
	The final dot is chosen so that the first row spans $\cN(\rT)$: this time $f_3=x^2+4y^2+4xy$ works. The result is a Jordan canonical basis and form for $\rT$:
	\[
		\beta=\bigl\{-4,\ \ 2y-x,\ \ xy,\ \ 4x+8y,\ \ x^2-4y^2,\ \ x^2+4y^2+4xy\bigr\},\quad\qquad 
		J=[\rT]_\beta=
		\left(
			\begin{array}{@{}c@{}c@{}c@{}}
				\begin{array}{@{\ }ccc@{\ }|}
					0 & 1 & 0\\
					0 & 0 & 1\\
					0 & 0 & 0\\\hline
				\end{array}&&\\[-0.4pt]
				&\begin{array}{|@{\ }cc@{\ }|}\hline
					0 & 1\\
					0 & 0\\\hline
				\end{array}&\\[-0.4pt]
				&&\begin{array}{|@{\ }c@{\ }}\hline
					0
				\end{array}
			\end{array}
		\right)
	\]
	As previously, other choices of cycle-generators $f_1,f_2,f_3$ are available. While alternatives result in different Jordan bases, Corollary \ref{cor:jordanunique} assures us that we'll always obtain the same canonical form $J$.
\end{example}


\pagebreak[3]


\begin{exercises}
	\hangindent\leftmargini\emph{Key concepts:\quad Dot diagrams: meaning \& computation,\quad Uniqueness of Jordan form}
  
	\begin{enumerate}
	  \item Let $\rT$ be a linear operator whose characteristic polynomial splits. Suppose the eigenvalues and the dot diagrams for the generalized eigenspaces $K_{\lambda_i}$ are as follows:
	  \[
	  	\begin{array}{c@{\qquad}c@{\qquad}c}
			  \lambda_1=2&\lambda_2=4&\lambda_3=-3\\
			  \begin{array}{ccc}
				  \bullet&\bullet&\bullet\\
				  \bullet&\bullet&\\
				  \bullet&&
			  \end{array}
			  &
			  \begin{array}{cc}
				  \bullet&\bullet\\
				  \bullet&\\
				  \bullet&
			  \end{array}
			  &
			  \begin{array}{cc}
				  \bullet&\bullet\\
				  &\\
				  &
		  	\end{array}
	  	\end{array}
	  \]
	  Describe the Jordan form $J$ of $\rT$.
  
  
		\item Suppose $\rT$ has Jordan canonical form $J$.\par
		\begin{minipage}[t]{0.6\linewidth}\vspace{0pt}
					\begin{enumerate}
		  \item State the characteristic polynomial of $\rT$.
		  \item For each eigenvalue $\lambda$, find the dot diagram and the smallest $k$ such that $K_{\lambda}=\cN(\rT-\lambda\I)^k$.
		\end{enumerate}
		\end{minipage}
		\hfill
		\begin{minipage}[t]{0.39\linewidth}\vspace{-25pt}
			\hfill
			$J=\left(
			\begin{array}{@{}c@{}c@{}c@{}c@{}}
				\begin{array}{@{\ }ccc@{\ }|}
					2 & 1 & 0\\
					0 & 2 & 1\\
					0 & 0 & 2\\\hline
				\end{array}&&&\\[-0.4pt]
				&\begin{array}{|@{\ }cc@{\ }|}\hline
					2 & 1\\
					0 & 2\\\hline
				\end{array}&&\\[-0.4pt]
				&&\begin{array}{|@{\ }c@{\ }|}\hline
					3\\\hline
				\end{array}&\\[-0.4pt]
				&&&\begin{array}{|@{\ }c@{\ }}\hline
					3
				\end{array}
			\end{array}
			\right)$
		\end{minipage}

		
		\item In Example \ref{ex:jordanpenult}, compute the two-cycle $\beta_{\vx_1}$ generated by $\vx_1=\ve_3$. Hence obtain an alternative Jordan canonical basis.
		
	  
		\item For each matrix $A$ find a Jordan canonical form and an invertible $Q$ such that $A=QJQ^{-1}$.
		\begin{enumerate}
		  \item $A=\begin{pmatrix}
		  -3 & 3 & -2\\
			-7 & 6 & -3\\
			1 & -1 & 2
		  \end{pmatrix}$\qquad
		  (b)\ \ $A=\begin{pmatrix}
		  0 & 1 & -1\\
			-4 & 4 & -2\\
			-2 & 1 & 1
		  \end{pmatrix}$\qquad
		  (c)\ \ $A=\begin{pmatrix}
		  0 & -3 & 1 & 2\\
			-2 & 1 & -1 & 2\\
			-2 & 1 & -1 & 2\\
			-2 & -3 & 1 & 4
		  \end{pmatrix}$
		\end{enumerate}
	
	
		\item For each linear operator $\rT$, find a Jordan canonical form $J$ and basis $\beta$:
		\begin{enumerate}
		  \item $\rT(f)=f'$ on $\Span_{\R}\big\{e^t,te^t,t^2e^{t},e^{2t}\big\}$
		  
		  \item $\rT\bigl(f(x)\bigr)=xf''(x)$ on $P_3(\R)$
		  
		  \item $\rT(f)=af_x+bf_y$ on $\Span_{\R}\big\{1,x,y,x^2,y^2,xy\big\}$. How does your answer depend on $a,b$?
		\end{enumerate}
		
		
		\item (Generalized Eigenvector Method for ODEs)\quad Let $A\in M_n(\R)$ have an eigenvalue $\lambda$ and suppose $\beta_{\vv_0}=\{\vv_{k-1},\ldots,\vv_1,\vv_0\}$ is a cycle of generalized eigenvectors for this eigenvalue.
		\begin{enumerate}
		  \item Show that $\vx(t):=e^{\lambda t}\sum\limits_{j=0}^{k-1}b_j(t)\vv_j$ is a solution to $\vx'(t)=A\vx$ if and only if
			\[
				\begin{cases}
					b_0'(t)=0,\text{ and}\\
					b'_j(t)=b_{j-1}(t)\text{ when $j\ge 1$}
				\end{cases}
			\]
			\item Use this method to find the general solution to the system of differential equations
			\[
				\vx'=\begin{pmatrix}
					3&1&0&0\\
					0&3&1&0\\
					0&0&3&0\\
					0&0&0&2
				\end{pmatrix}
				\vx
			\]
		\end{enumerate}
	\end{enumerate}
\end{exercises}



\clearpage


 
\subsection{The Rational Canonical Form (non-examinable)}

We finish with a very quick discussion of what is possible when the characteristic polynomial of a linear map does not split. In such a situation, we may assume that
\[
	p(t)=(-1)^n\bigl(\phi_1(t)\bigr)^{m_1}\cdots\bigl(\phi_k(t)\bigr)^{m_k}\tag{$\ast$}
\]
where each $\phi_j(t)$ is an \emph{irreducible monic polynomial} over the field.

\begin{example}{}{ratcon1}
	The following matrix has characteristic equation $p(t)=(t^2+1)^2(3-t)$:
	\[
		A=
		\begin{pmatrix}
			0 & -1 & 0 & 0 & 0\\
			1 & 0 & 0 & 0 & 0\\
			0 & 0 & 0 & -1 & 0\\
			0 & 0 & 1 & 0 & 0\\
			0 & 0 & 0 & 0 & 3
		\end{pmatrix}
		\in M_5(\R)
	\]
	This doesn't split over $\R$ since $t^2+1=0$ has no real roots. It is, however, diagonalizable over $\C$.
\end{example}

A couple of basic facts from algebra:
\begin{itemize}
  \item Every polynomial splits over $\C$: every $A\in M_n(\C)$ therefore has a Jordan form.
  \item Every polynomial over $\R$ factorizes into linear or irreducible quadratic factors.
\end{itemize}

The question is how to deal with non-linear irreducible factors in the characteristic polynomial.

\begin{defn}{}{}
	A monic polynomial $t^k+a_{k-1}t^{k-1}+\cdots+a_0$ has \emph{companion matrix}
	\[
		\begin{pmatrix}
			0 & 0 & 0 & \cdots & 0 & -a_0\\
			1 & 0 & 0 & & 0 & -a_1\\
			0 & 1 & 0 & & 0 & -a_2\\
			\vdots & & & \ddots & & \vdots\\
			0 & 0 & 0 & & 0 & -a_{k-2}\\
			0 & 0 & 0 & \cdots & 1 & -a_{k-1}
		\end{pmatrix}
		\tag{when $k=1$, this is the $1\times 1$ matrix $(-a_0)$}
	\]
	If $\rT\in\cL(V)$ has characteristic polynomial ($\ast$), then a \emph{rational canonical basis} is a basis for which
	\[
		[\rT]_\beta=
		\begin{pmatrix}
			C_1&O&\cdots&O\\
			O&C_2&&O\\
			\vdots&&\ddots&\vdots\\
			O&O&\cdots&C_r
		\end{pmatrix}
	\]
	where each $C_j$ is a companion matrix of some $(\phi_j(t))^{s_j}$, $s_j\le m_j$. We call $[\rT]_\beta$ a \emph{rational canonical form}.
\end{defn}

The matrix in Example \ref{ex:ratcon1} is a rational canonical form and the standard basis is rational canonical. We state the main result without proof.

\begin{thm}{}{}
	A rational canonical basis exists for \emph{any} linear operator $\rT$ on a finite-dimensional vector space $V$. The canonical form is unique up to ordering of companion matrices.
\end{thm}


\begin{example}{}{}
	$A=
	\begin{smatrix}
		4&-3\\
		2&2
	\end{smatrix}
	\in M_2(\R)$  characteristic polynomial
	\[
		p(t)=t^2-6t+14=(t-3)^2+5
	\]
	This doesn't factorize over $\R$, so it has no real eigenvalues. Instead simply pick a vector, $\vx=\stwovec 10$ (say), define $\vy=A\vx=\stwovec 42$, let $\beta=\{\vx,\vy\}$ and observe that
	\[
		[\rL_A]_\beta=
		\begin{pmatrix}
			0&-14\\
			1&6
		\end{pmatrix}
	\]
	is a rational canonical form. Indeed this works for \emph{any} $\vx\neq\V0$: if $\beta:=\{\vx,A\vx\}$, then Cayley--Hamilton forces
	\[
		A^2\vx=(6A-14I)\vx=-14\vx+6A\vx
		\implies  [\rL_A]_\beta=
		\begin{pmatrix}
			0&-14\\
			1&6
		\end{pmatrix}
	\]
	Otherwise said, $\beta$ is a rational canonical basis and $[\rL_A]_\beta$ is independent of $\vx$!
\end{example}

A systematic approach to finding rational canonical forms is similar to that for Jordan forms: for each irreducible divisor of $p(t)$, the subspace $K_\phi=\cN\bigl(\phi(\rT)\bigr)^m$ plays a role analogous to a generalized eigenspace; indeed $K_\lambda=K_\phi$ for the \emph{linear} irreducible factor $\phi(t)=\lambda-t$.

\begin{examples}{}{}
	Suppose a $4\times 4$ matrix has characteristic polynomial
	\[
		p(t)=\big(\phi(t)\big)^2
		=(t^2-2t+3)^2
		=t^4-4t^3+10t^2-12t+9
	\]
	There are two possible rational canonical forms corresponding to this polynomial, we provide an example of each.
Hopefully the approach is intuitive, even without theoretical justification.
	\begin{enumerate}
	  \item If $A=
	  \begin{smatrix}
			  0&-15&0&-9\\
			  2&2&-3&0\\
			  0&-9&0&-6\\
			  -3&0&5&2
	  \end{smatrix}$,
	  then $\phi(A)=O$ is the zero matrix, whence $\cN(\phi(A))=\R^4$. Since $\phi(t)$ isn't the full characteristic polynomial, we expect there to be \emph{two} independent cycles of length two in the canonical basis. Start with something simple as a guess:
	  \[
	  	\vx_1=\sfourvec 1000\implies \vx_2=A\vx_1=\sfourvec 020{-3}
	  	\implies A\vx_2=\sfourvec{-3}40{-6}=-3\vx_1+2\vx_2
	  \]
	  Now make another choice that isn't in the span of $\{\vx_1,\vx_2\}$:
	  \[
	  	\vx_3=\sfourvec 0010\implies \vx_4=A\vx_3=\sfourvec 0{-3}05
	  	\implies A\vx_4=\sfourvec 0{-6}{-3}{10}=-3\vx_3+2\vx_4
	  \]
	  We therefore have a rational canonical basis $\beta=\{\vx_1,\vx_2,\vx_3,\vx_4\}$ and can express $A$ more simply as
	  \[
		  A=
		  \begin{smatrix}
			  1&0&0&0\\
			  0&2&0&-3\\
			  0&0&1&0\\
			  0&-3&0&5
		  \end{smatrix}
		  \begin{smatrix}
			  0&-3&0&0\\
			  1&2&0&0\\
			  0&0&0&-3\\
			  0&0&1&2
		  \end{smatrix}
		  \begin{smatrix}
			  1&0&0&0\\
			  0&2&0&-3\\
			  0&0&1&0\\
			  0&-3&0&5
		  \end{smatrix}
		  ^{-1}
	  \]
	  Over $\C$, this example is diagonalizable. Indeed each of the $2\times 2$ companion matrices is diagonalizable over $\C$, with eigenvalues $\lambda=1\pm\sqrt 2i$.
	  
	  \item Let $B=
	  \begin{smatrix}
		  0&0&2&1\\
		  1&1&-1&-1\\
		  0&1&-2&-16\\
		  0&0&1&5
	  \end{smatrix}$. 
	  This time
	  \[
	  	\phi(B)=B^2-2B+3I=
	  	\begin{smatrix}
			  3&2&-7&-29\\
			  -1&1&4&13\\
			  1&-3&-6&-17\\
			  0&1&1&2
	  	\end{smatrix}
	  	\implies \cN(\phi(B))=\Span\left\{\sfourvec 3{-1}10,\sfourvec{11}{-2}01\right\}
	  \]
	  Anything not in this span will suffice as a generator for a single cycle of length four: e.g.,
	  \begin{gather*}
		  \vx_1=\sfourvec 1000,\quad \vx_2=B\vx_1=\sfourvec 0100,\quad 
		  \vx_3=B\vx_2=\sfourvec 0110,\quad \vx_4=B\vx_3= \sfourvec 20{-1}1\\
		  B\vx_4=\sfourvec{-1}2{-14}4=-9\sfourvec 1000+12\sfourvec 0100-10\sfourvec 0110+4\sfourvec 20{-1}1
	  \end{gather*}
	  We therefore have a rational canonical basis $\beta=\{\vx_1,\vx_2,\vx_3,\vx_4\}$ and
	  \[
	  	B=
		  \begin{smatrix}
			  1&0&0&2\\
			  0&1&1&0\\
			  0&0&1&-1\\
			  0&0&0&1
		  \end{smatrix}
		  \begin{smatrix}
			  0&0&0&-9\\
			  1&0&0&12\\
			  0&1&0&-10\\
			  0&0&1&4
		  \end{smatrix}
		  \begin{smatrix}
			  1&0&0&2\\
			  0&1&1&0\\
			  0&0&1&-1\\
			  0&0&0&1
		  \end{smatrix}
		  ^{-1}
		 \]
	  In contrast to the first example, $B$ isn't diagonalizable over $\C$. It has two eigenvalues $\lambda=1+i\sqrt 2$ and its conjugate $\cl\lambda=1-\sqrt 2i$$\lambda=1+i\sqrt 2$, each of multiplicity two, and Jordan form
	  \[
	  	J=\begin{smatrix}
			  \lambda&1&0&0\\
			  0&\lambda&0&0\\
			  0&0&\cl\lambda&1\\
			  0&0&0&\cl\lambda
	  	\end{smatrix}
	  \]
	\end{enumerate}
\end{examples}

